Gammapy: A Python package for gamma-ray astronomy

### 2023-01-20 src/ms-review.tex
Apply suggestions from code review
Co-authored-by: Cosimo Nigro <cosimonigro2@gmail.com>
Co-authored-by: Bruno Khélifi <khelifi@in2p3.fr>

\abstract{
	% TODO: the feature are still to detailed and partly copy and pasted form the ICRS abstract...avoid this
	% context heading (optional)
	% {} leave it empty if necessary
	%{Gammapy context:}
	Historically, the data and the analysis software of \gammaray astronomy experiments are
	proprietary to the collaborations operating them. The next generation of \gammaray instruments, as the Cherenkov Telescope Array
	(CTA) will be operated as open observatories, with public
	data. This introduces the need to have a high-level analysis software available to the astronomical community. In
	this article we present the first major release of \gammapy, the Long-term Support (LTS) v1.0, a
	community-developed open-source Python package for \gammaray astronomy. We
	present its general design and provide an
	overview of the analysis methods and features it implements. Starting from
	event lists and a description of the specific instrument response functions
	(IRF) stored in open FITS based data formats, \gammapy implements the reduction
	of the input data and instrument response to binned WCS, HEALPix or region based
	data structures. Thereby it
	handles the dependency of the IRFs with time, energy as well as position on the
	sky. It offers a variety of background estimation methods for spectral, spatial
	and spectro-morphological analysis. Counts, background and IRFs data are
	bundled in datasets and can be serialised, rebinned and stacked. \gammapy
	supports to model binned data using Poisson
	maximum likelihood fitting. It comes with built-in spectral, spatial and
	temporal models, support for particle radiation emission spectra as well as
	support for custom user models, to model e.g.,
	energy dependent morphology of \gammaray sources. Multiple datasets can be
	combined in a joint-likelihood approach to either handle time dependent IRFs,
	different classes of events or combination of data from multiple instruments.
	Gammapy also implements methods to estimate flux points, likelihood
	profiles per energy bin, light curves as well as flux and signficance maps in
	energy bins.  We further describe the general
	development approach and how \gammapy integrates into ecosystem of other
	scientific and astronomical Python packages. We also present analysis examples
	with simulated CTA data and provide results of scientific validation analyses
	using data of existing instruments such as \hess and \fermi.

	% The \gammaray can be considered as the last frontier
	% In the last two decades the measurement of \gammaray emission from the universe has
	% evolved from a niche

	% aims heading (mandatory)
	%{Gammapy aims}
	%Reproducible analyses, share algorithms etc.

	% methods heading (mandatory)
	%{Gammapy methods}

	% results heading (mandatory)
	%{Gammapy results}

	% conclusions heading (optional), leave it empty if necessary
	%{Gammapy conclusions}
}

@@

The other sequence, referred to as high-level analysis, deals with the extraction of physical
quantities related to \gammaray sources and the production of high-level products such as spectra,
light curves and catalogs. The methods applied here are more generic and are broadly
shared across the field. They also frequently imply joint analysis of multi-instrument data.
To extract physically relevant information, the measured counts on an instrument are usually
compared to a parametrised model of the true intensity distribution of \gammaray source in a given
field-of-view, which is "forward folded" through a model of the instrument response
to get a prediction for the expected counts measured by the instrument. Parameters
of the model are then estimated using statistical techniques such as maximum likelihood
estimation (MLE).

\subsection{Predicted counts and Instrument Response Functions}

In the most general way we can write the expected number of detected events from a given astrophysical source
of \gammaray at measured position $p$ and energy $E$:
\begin{align}
   N(p, E) {\rm d}p {\rm d}E = &t_{\rm obs} \int_{E_{\rm true}} \int_{p_{\rm true}}  R(p, E|p_{\rm true}, E_{\rm true})\\
   &\times \Phi(p_{\rm true}, E_{\rm true}) {\rm d}E_{\rm true} {\rm d}p_{\rm true}
\end{align}
where
\begin{itemize}
\item $R(p, E| p_{\rm true}, E_{\rm true})$ is the instrument response
\item $\Phi(p_{\rm true}, E_{\rm true})$ is the parametrised sky flux model
\item $t_{\rm obs}$ is the observation time
\end{itemize}

@@

\gammaray data at the Data Level 3 (DL3) therefore consist of lists of gamma-like events and their
corresponding instrument response functions (IRFs). The latter include the effective area ($A_{\rm eff}$),
point spread function (PSF) and energy dispersion ($E_{\rm disp}$).

An additional component of DL3 IRFs is the residual hadronic background model $Bkg$.
It represents the intensity of charged particles misidentified as gamma-rays that are expected
during an observation. It is defined as a function of the measured position in the field-of-view
and measured energy.

In total, the expected number of events in a \gammaray observation is given by:
\begin{align}
N(p, E) {\rm d}p {\rm d}E =  E_{\rm disp} \star (PSF \star A_{\rm eff} t_{\rm obs} \Phi) + Bkg(p, E)
\end{align}

In practice, the background model is subject to numerous uncertainties and needs to be corrected on the data.
More frequently, a background model is not even available and the background contribution has to be estimated
directly from the data.

\subsection{Data analysis workflow}

In Gammapy, the handling of DL3 data is performed by classes and methods
in the \code{gammapy.data} (see Section~\ref{ssec:gammapy-data}) and the \code{gammapy.irf}
(see Section~\ref{ssec:gammapy-irf}) subpackages.

@@

The next step of the analysis is the data reduction, where all observation events and instrument
responses are filled into or projected onto a common physical coordinate system, defined by
a map geometry. The definition of the map geometry typically consists of a spectral dimension
defined by a binned energy axis and of spatial dimensions, which either define 
a spherical projection from celestial coordinates to a pixelised image space
(for 3-dimensional or cube analysis) or a region on the sky (for regular spectral analysis).
The \code{gammapy.maps} subpackage provides general multidimensional geometry objects
(\code{Geom}) and the associated data structures (\code{Maps}) (see Section~\ref{ssec:gammapy-maps}).

After all data has been projected into the same geometry, it is typically
required to improve the residual hadronic background estimate. As residual hadronic
background models can be subject to significant systematic uncertainties,
the background estimated can be improved by taking into account actual data
from regions without known \gammaray sources. This includes methods 
such as the ring or the field-of-view background techniques or
background measurements performed within, e.g. reflected regions~\citep{Berge07}.
Data measured at the field-of-view or energy boundaries of the instrument is typically
associated with a systematic uncertainty in the IRF. For this reason this part 
of the data is often excluded from subsequent analysis by defining regions of
 "safe" data in the spatial as well as energy dimension.
All of these data reduction steps are performed by classes and functions
implemented in the \code{gammapy.makers} subpackage (see Section~\ref{ssec:gammapy-makers}).

The counts data and the reduced IRFs in the form of maps are bundled into datasets
that represent the data level 4 (DL4). Those reduced datasets can be written to disk,
in a format specific to \gammapy to allow users to read them back at any time later
for modeling and fitting. Different variations of such datasets support different 
analysis methods and fit statistics (e.g. Poisson statistics with known background or
with OFF background measurements). The datasets are used to perform joint-likelihood
fitting allowing to combine different measurements, e.g. from different observations
but also from different instruments or event classes. They can also be used for binned
simulation as well as event sampling to simulate DL3 events data.
The various DL4 objects and the associated functionalities are
implemented in the \code{gammapy.datasets} subpackage (see Section~\ref{ssec:gammapy-datasets}).

The next step is then typically to model and fit the datasets, either
individually, or in a joint likelihood analysis. For this purpose \gammapy
provides a uniform interface to multiple fitting backends. It also provides
a variety of built in models. This includes spectral,
spatial and temporal model classes to describe the \gammaray emission in the sky.
Where spectral models can be simple analytical models or more complex ones from radiation
mechanisms of accelerated particle populations (e.g. inverse Compton or $\pi^{o}$ decay).
Independently or subsequently to the global modelling, the data can be
re-grouped to compute flux points, light curves and flux as well as significance
maps in energy bands.
The modelling and fitting functionalities are implemented in the \code{gammapy.modeling},
\code{gammapy.estimators} and \code{gammapy.stats} subpackages (see respectively
Section~\ref{ssec:gammapy-modeling}, \ref{ssec:gammapy-estimators} and \ref{ssec:gammapy-stats}).

>>>>

\abstract{

	{Context} Traditionally, the \gammaray universe has been probed
	 by small collaborations of scientists employing proprietary data and analysis software.
	 However, the next generation of \gammaray instruments,
	 such as the the Cherenkov Telescope Array (CTA), will be operated as open observatories.
	 Alongside the data, the observatories will make available to the astronomical community
	 software tools for their analysis. This necessity prompted the development, in the last decade,
	 of open high-level astronomy software customised for high energy astrophysics.

	{Aims} In this article, we present \gammapy, an open-source python package for the analysis of \gammaray data,
	and illustrate the functionalities of the first long-term release of the software, version 1.0.
	Built on the modern python scientific ecosystem, \gammapy provides a uniform platform for reducing and
	modelling data from different gamma-ray instruments. \gammapy complies with several well-established
	data conventions in high-energy astrophysics, providing serialised data products that are interoperable
	with other software.

	{Methods} Starting from event list and instrument response functions",
	\gammapy provides pipelines for creating reduced data binned in energy and coordinates.
	To facilitate the handling of the hadronic background, several techniques for background estimation
	 are implemented in the package.
	 After the data are binned, the flux and morphology of one or more \gammaray sources can be estimated
	 using Poisson maximum likelihood fitting
	 and assuming a variety of spectral, temporal, and spatial models.
	  Estimation of flux points, likelihood profiles and light curves are also implemented.

	{Results} After describing the structure of the package, we demonstrate the capabilities of \gammapy
	performing several traditional analyses in gamma-ray astronomy with publicly available data:
	spectral and spectro-morphological analyses, estimation of the light curve of a source.
	The flexibility and power of the software are displayed in a final multi-instrument example,
	where data sets from different instruments, at different stages of data reduction,
	are simultaneously fitted with a flux model produced by another astrophysical software.
}

@@

The other sequence, referred to as high-level analysis, deals with the extraction of physical
quantities related to \gammaray sources and the production of high-level products such as spectra,
light curves and catalogs. The methods applied here are more generic and are broadly
shared across the field. They also frequently imply joint analysis of multi-instrument data.
To extract physically relevant information, the measured counts on an instrument are usually
compared to a parametrised model of the true intensity distribution of \gammaray source in a given
field-of-view, which is "forward folded" through a model of the instrument repsonse
to get a prediction for the expected counts measured by the instrument. Parameters
of the model are then estimated using statistical techniques such as maximum likelihood
estimation (MLE).

In the most general way we can write the expected number of detected events at measured
position $p$ and energy $E$:
\begin{align}
   N(p, E) {\rm d}p {\rm d}E = &t_{\rm obs} \int_{E_{\rm true}} \int_{p_{\rm true}}  R(p, E|p_{\rm true}, E_{\rm true})\\
   &\times \Phi(p_{\rm true}, E_{\rm true}) {\rm d}E_{\rm true} {\rm d}p_{\rm true}
\end{align}
where
\begin{itemize}
\item $R(p, E| p_{\rm true}, E_{\rm true})$ is the instrument response
\item $\Phi(p_{\rm true}, E_{\rm true})$ is the parametrised sky flux model
\item $t_{\rm obs}$ is the observation time
\end{itemize}

@@

\gammaray data at the Data Level 3 (DL3) therefore consists in lists of gamma-like events and their
corresponding instrument response functions (IRFs). The latter include the aforementioned
effective area, point spread function (PSF), energy dispersion and residual hadronic background.
The handling of DL3 data is performed by classes and methods
in the \code{gammapy.data} (see Section~\ref{ssec:gammapy-data}) and the \code{gammapy.irf}
(see Section~\ref{ssec:gammapy-irf}) subpackages.

@@

The next step of the analysis is the data reduction, where all observation events and instrument
responses are filled into or projected onto a common physical coordinate system, defined by
a map geometry. The definition of the map geometry typically consists of a spectral dimension
defined by a binned energy axis and of a spatial dimension, which either defines 
a spherical projection from celestial coordinates to an pixelised image space
(for 3-dimensional or cube analysis) or a region on the sky (for regular spectral analysis).
The \code{gammapy.maps} subpackage provides general multidimensional geometry objects
(\code{Geom}) and the associated data structures (\code{Maps}) (see Section~\ref{ssec:gammapy-maps}).

After all data has been projected into the same coordinate system, it is typically
required to improve the residual hadronic background estimate. As residual hadronic
background models can be subject to significant systematic uncertainties,
the background estimated can be improved by taking into account actual data
from regions without known \gammaray sources. This includes methods 
such as the ring or the field-of-view background techniques or
background measurements performed within, e.g. reflected regions~\citep{Berge07}.
Data measured at the field-of-view or energy boundaries of the instrument is typically
associated with a sytematic uncertainty in the IRF. For this reason this part 
of the data is often excluded from subsequent analysis by defining regions of
 "safe" data in the spatial as well as energy dimension.
All of these data reduction steps are performed by classes and functions
implemented in the \code{gammapy.makers} subpackage (see Section~\ref{ssec:gammapy-makers}).

The counts data and the reduced IRFs in the form of maps are bundled into datasets
that represent the data level 4 (DL4). Those reduced datasets can be written to disk,
in a format specific to \gammapy to allow users to read them back at any time later
for modeling and fitting. Different variations of such datasets support different 
analysis methods and fit statistics (e.g. Poisson statistics with known background or
with OFF background measurements). The datasets are used to perform joint-likelihood
fitting allowing to combine different measurements, e.g. from different observations
but also from different instruments or event classes. They can also be used for binned
simulation as well as event sampling to simulate DL3 events data.

The next step is then typically to model and fit the datasets, either
individually, or in a joint likelihood analysis. For this purpose \gammapy
provides a uniform interface to multiple fitting backends. It also provides
a variety of built in models. This includes spectral,
spatial and temporal model classes to describe the \gammaray emission in the sky.
Where spectral models can be simple analytical models or more complex ones from radiation
mechanisms of accelerated particle populations (e.g. inverse Compton or $\pi^{o}$ decay).
Independently or subsequently to the global modelling, the data can be
re-grouped to compute flux points, light curves and flux as well as significance
maps in energy bands.
